---
title: "RRweek2 Assignment"
author: "Andy"
date: "4/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Hey guys this is my first R markdown document! Take a look at my work below!

## Part 1 - Loading and Processing the data.
```{r getdata, echo=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "activitydata.zip") #downloads data from coursera provided link.
unzip("activitydata.zip") #unzips the data in the zip file
rawdata  <- read.table("activity.csv", sep = ",", header = TRUE) #since header is provided in the .csv file I've included it as true and denoted the seperator as a comma. (you can use read.csv as well)
str(rawdata) #this will show you a some stats about the data.
library(ggplot2)
library(dplyr)
library(lubridate)
```

You can see from the summary data above, that there are 17,568 observations under 3 variables. Not at this time that the dat is in class "factor" format.

## Part 2 - What is mean total number of steps taken per day?

```{r totalstepsperday, echo=TRUE}
totstepsbyday <- tapply(rawdata$steps, rawdata$date, sum, na.rm = TRUE) #outputs a list/array of values while ignoring NA values
hist(totstepsbyday,bins= 100, main = "Total Steps by Day (histogram)", xlab = "Total Steps") #outputs the frequency of total steps by day
print(c("The mean of the total steps is:",mean(totstepsbyday),"The Median of the total steps is:", median(totstepsbyday)))
```
## Part 3 - What is the average daily activity pattern?

```{r avgdailyactivity, echo=TRUE}
avgactivity <- rawdata %>% group_by(interval) %>% summarize(avgstep = mean(steps,na.rm=TRUE))  #calcuates total steps taken at each inverval across all days
qplot(interval, avgstep, data = avgactivity, geom = "line", xlab = "5 minute intervals", ylab = "average steps taken across all days)")
highestinterval<- avgactivity[which.max(avgactivity$avgstep),1]
print(c("highest average step occurs in interval:", highestinterval))

```


## Part 4 - Imputing missing values & recaluculate missing days

```{r imputemissingvalues, echo=TRUE}
misvalues <- rawdata[is.na(rawdata),] #subsets the missingvalues data.
print(c(nrow(misvalues), "rows of missing data"))

imputedata <- merge(rawdata, avgactivity)  #means data was calculated in last step - I can re use the data per interval here.
imputedata$steps[is.na(imputedata$steps)] <- imputedata$avgstep[is.na(imputedata$steps)]
dataforhits <- group_by(imputedata, date) %>% summarize(steps = sum(steps))
qplot(steps, data = dataforhits, main = "Histogram of Total Steps by Day with Imputed Data")
print(c("The mean total steps/day is", mean(dataforhits$steps)," The median total steps/day is", median(dataforhits$steps)))
```

## Part 5 - Differences in weekdays vs weekends!
```{r weekdayvsweekend, echo=TRUE}
imputedata$date <- ymd(imputedata$date)  
weekdaydata <- mutate(imputedata, whichday = weekdays(imputedata$date))
weekdaydata <- mutate(weekdaydata, weekend =  as.factor(ifelse(weekdaydata$whichday == "Saturday" | weekdaydata$whichday == "Sunday","weekend", "weekday")))
meansteps <- weekdaydata %>% group_by(interval,weekend) %>%   summarise(average = mean(steps))
qplot(interval, average, data = meansteps, facets = weekend~., geom = "line", main = " Average Number of Steps between Wkday and Wkend", xlab = "Intervals (5 minute)", ylab = "AVG number of Steps Taken")
```